Constraining the Interiors of Asteroids Through Close Encounters

### 2022-08-26 paper/main.tex
Edited paper

\begin{abstract}
  Knowledge of the interior density distribution of an asteroid can reveal its composition and constrain its evolutionary history. However, most asteroid observational techniques are not sensitive to interior properties. We investigate the interior constraints accessible through monitoring variations in angular velocity during a close encounter. We derive the equations of motion for a rigid asteroid's orientation and angular velocity to arbitrary order and use them to generate synthetic angular velocity data for a representative asteroid on a close Earth encounter. Using Markov Chain Monte Carlo fits, we perform injection-retrieval tests on these synthetic data to gain insights into the extent to which interior properties can be constrained. We also perform a sensitivity analysis of such an inversion technique to asteroid parameters (e.g., asteroid shape and orbital elements), observational set-up (e.g., measurement precision and cadence), and mapping models to convert constraints on the density moments to density distributions. We find that high precision in rotational period estimates (order of fifty milliseconds) are necessary for each cadence, and that low perigees ($\lesssim 10$ Earth radii) are necessary to resolve density distributions.
\end{abstract}

@@

We begin our simulation at $D = 10 r_p$. Since the leading order of equation \ref{eqn:tidal-torque} yields $\tau \propto D^{-3}$, this corresponds roughly to a torque of $10^{-3}$ times the maximum torque at perigee. Unless otherwise indicated, the simulation is terminated at $D=10 r_p$ as well. With the simulation inputs specified, the equations of motion are integrated via the Runge-Kutta fourth order method, with a variable time step, manually chosen to limit integration error to only \~100 times the floating point numerical error.

@@

Before the MCMC was run, local minima in the likelihood were found via the Nelder-Mead algorithm implemented in \texttt{scipy} \citep{Gao2012}. Generally, only one local minimum existed, except when $K_{22}=0$ in which case rotational symmetry caused all values of $\gamma_0$ to be degenerate. Ensemble walkers were initialized near the local minimum, distributed by a Gaussian approximation of the likelihood, as determined via the Hessian of the likelihood at the minimum. Due to the high sensitivity of the angular velocity data to density moments, the minimization procedure sometimes failed to isolate the minimum likelihood. Therefore, a simpler simulation without the $K_{3m}$ terms of equation \ref{eqn:tidal-torque} was first used to minimize likelihood as a function of the first-order parameters $\gamma_0$ and $K_{2m}$, and then the full simulation was used to find the second-order parameters $K_{3m}$ with the first-order parameters fixed.

@@

We assume that the asteroid's surface is known in the inertial frame from radar data. Since the asteroid tumbles during the encounter, we also assume that the center of mass of the asteroid is extracted in the inertial frame. The density moments are extracted from flyby data, but these are known instead in the body-fixed frame.

To compare the asteroid surface and extracted density moments in the same coordinate system, we define a new frame called the ``hybrid frame.'' The hybrid frame is co-located with the body-fixed frame, but its orientation is known with respect to the inertial frame; it has $\bm \Omega_0 \parallel \unit z_\text{hybrid}$ with third Euler angle $\gamma = 0$. The hybrid frame differs from the body-fixed frame only by a rotation around $\unit z_\text{hybrid}=\unit z_\text{body-fixed}$ of $\gamma_0$. Such rotations affect density moments via 
\begin{equation}
  K_{\ell m}^\mathrm{hybrid} = e^{-im\gamma_0}K_{\ell m}^\mathrm{body-fixed}.
  \label{eqn:body-fixed-to-hybrid}
\end{equation}
by equation \ref{eqn:ylm-rotation}. Thus, values and  uncertainties on $K_{\ell m}^\mathrm{body-fixed}$ and $\gamma_0$ (obtained from the encounter data) can be translated into values and uncertainties on $K_{\ell m}^\mathrm{hybrid}$. The surface model can be rotated from the inertial frame to the hybrid frame so that both the surface and $K_{\ell m}$ are now in the same frame and are comparable. Henceforth, we will operate only in the hybrid frame and suppress the label.

@@

To define the ``finite element'' model, we divide the asteroid into $N$ finite elements of uniform density and use the density $\rho_i$ of each as parameters. The four constraints imposed by the known mass and center of mass of the asteroid (seven with $K_{21}$ and $\Im K_{22}$ are also fixed) are used to fix some of the $\rho_i$. These constrained densities are easily computed since the asteroid mass $\mu_\mathcal{A}$ and the product $I_\mathcal{A}K_{\ell m}$ are both linear functions of $\rho_i$. Because $I_\mathcal{A}K_{00} = a_\mathcal{A}^2\mu_\mathcal{A}$ by definition (which is known), and $K_{\ell m} = 0$ implies that $I_\mathcal{A}K_{\ell m} = 0$, $I_\mathcal{A} K_{\ell m}$ is known for all the fixed moments, and computing the corresponding constrained densities is simply a matrix inversion. The size and location of each finite element must be chosen before extracting its density.

@@

The translation rules of spherical harmonics \citep{Gelderen1998TheSO} give that the density moments in the hybrid frame are
\begin{equation}
  \begin{aligned}
    K_{\ell m} = &\brackets{\sum_{i=0}^N \frac{a_i^2 + r_i^2}
    {a_\mathcal{A}^2}\mu_i}^{-1} \\
    &\times \brackets{\sum_{i=0}^N \sum_{\ell' m'}\mu_i
    \frac{a_i^{\ell'}}{a_\mathcal{A}^\ell}
    R_{\ell - \ell', m - m'}(\bm r_i)K_{\ell' m'}^{(i)}} \\
  \end{aligned}
\end{equation}
where the unmarked sum limits are $0 \leq l' \leq l$ and $-\ell' \leq m' \leq \ell'$. We also have total mass and center of mass constraints:
\begin{equation}
  \mu_\mathcal{A} = \sum_{i=0}^N \mu_i;  \qquad 0 = \sum_{i=0}^N \mu_i \bm r_i.
\end{equation}
Additional constraints can be imposed on $K_{\ell m}^{(i)}$ if desired. For example, we can require that the lumps be ellipsoids, so that $K_{3 m}^{(i)} = 0$. The most extreme case is spherical lumps, which have $K_{\ell m} = 0$ for $\ell > 0$. $K_{00}=1$ is also guaranteed by definition, meaning that each spherical lump has only five DOF ($a_i$, $\mu_i$, and $\bm r_i$). The substrate has one degree of freedom, since only $\mu_0$ is unknown. Thus, this spherical lumpy model has $5N - 3$ total DOF. Again, the choice of $N$ affects the accuracy and uncertainty of the model results. For the rest of this paper, we use $N=1$ spherical lump for simplicity, corresponding to 2 DOF. In section \ref{sec:density-compare}, we also use $N=2$ spherical lumps which possesses 7 DOF.

@@

Once the parameters of the density distribution models have been extracted, each MCMC sample corresponds to its own density distribution. To generate an average distribution, we randomly choose 1,000 of these samples and define the density distribution $\rho$ at each point to be the mean of the densities at that point across the 1,000 samples, and take the standard deviation of the sample densities as the density uncertainty $\sigma_\rho$. In the case of the finite element model, we run the MCMC twenty times and select 1,000 parameter samples from each to reduce dependence on the initial choice of finite element locations.

@@

\begin{enumerate}
  \item An orbit around a spherical, non-rotating, Moonless Earth with $6$ km s$^{-1}$ excess velocity and perigee at 5 Earth radii. This orbit was chosen to roughly match that of Apophis and corresponds to an eccentricity of 3.88. 
  \item An initial roll of $\gamma_0=\pi/8$.
  \item A cadence of 2 minutes and observational uncertainty of $\sigma_\theta = 10^{-2}$ and $\sigma_P / P = 10^{-7}$.
  \item A rotational period of 9 hours, with the initial angular velocity vector distributed between the $\unit X$, $\unit Y$, and $\unit Z$ axes in a $1:2:-2$ ratio.
  \item An asteroid with radius $a_\mathcal{A} = 1$ km and $K_{3m}=0$. For $K_{22}$ and $K_{20}$, we use two standard values: one with $(K_{22}, K_{20}) = (0, -0.097)$ and one with $(0.052, -0.202)$. Including the third point obtained by reflection $K_{22}\rightarrow -K_{22}$ (a 90$^\circ$ rotation), these are the three points that minimize the mean distance between an arbitrary point in the allowed parameter space (equation \ref{eqn:parameter-bounds}) and these reference values. The first point is called the symmetric case because the corresponding uniform-density-ellipsoid model is rotationally symmetric around $\unit z$. The second case and its reflection are called the asymmetric cases. Values of $(0.052, -0.202)$ have $a < b$ in the ellipsoid model, and the reflected value has $a > b$. Specifically, the asymmetric case has $a=1140$ m, $b=1839$ m, and $c=565$ m, while the symmetric case has $a=b=1411$ m and $c=1008$ m.
\end{enumerate}

In the following sections, we first introduce the retrieval capabilities regarding density moments, then we turn to an example of a retrieval of the density distribution from the moments. For both stages of information retrieval, we find that the results are consistent with the properties used to generate the synthetic data.

\subsection{Density moment retrieval}
\label{sec:results-moment}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/example-residuals.pdf}
  \caption{Data, best-fitting results, and residuals for a fit to synthetic data simulated for an asymmetric reference asteroid. Uncertainty bands are also shown. The best fit results are consistent with the data.}
  \label{fig:example-residuals}
\end{figure}

@@

Figure \ref{fig:example-corner} shows a corner plot of the PPDs of the ten parameters (namely $\gamma_0$ and $K_{\ell m}$ for $\ell \leq 3$), marginalized to functions of one (histograms) or two (contours) variables. The true parameters are shown and usually lie within 1 or 2$\sigma$ of the $\Delta K_{\ell m} = 0$, where $\Delta K_{\ell m}$ is the difference between the posterior $K_{\ell m}$ and the true $K_{\ell m}$. The PPDs are generally Gaussian and sometimes show strong correlation between parameters, but no continuous degeneracy occurs. We performed 48 independent minimizations of the likelihood before the MCMC fit began, each with an initial point chosen randomly in the parameter space. All converged to the same minimum, demonstrating that the model lacks discrete degeneracy as well.

@@

To provide an example of a density distribution extraction, we consider an asteroid with a core and ask whether our methodology can resolve the location, mass, and size of the core. Specifically, we use a spherical core of radius 300 m, placed 500 m from the center and with density 50\% greater that of the surrounding asteroid. We use the lumpy model to extract a distribution in this section because it is designed to look for cores.

Synthetic data was generated for the new asteroid and moments were extracted via the process described above. These moments were consistent with the true moments, with a $\chi^2$ per DOF of $\chi^2_r = 1.1$, and inconsistent with the moments of a nearly-uniform asteroid of the same shape ($\chi^2_r = 2.7 \times 10^8$). The inconsistency comes almost entirely from the MOI-constraining $K_{2m}$ moments, which are sensitive to large mass distribution changes such as the presence of a core.

@@

  \includegraphics[width=0.49\linewidth]{figs/side-by-side-u}
  \caption{The extracted distribution (\textit{left}) and true distribution (\textit{right}) of the asymmetric reference asteroid. The density distribution was created via the lumpy model. Uncertainty in the distribution is shown below. The extracted distribution is nearly identical to the true distribution and is statistically consistent with the truth.}
  \label{fig:den-side-by-side}
\end{figure}

A density distribution was then extracted, shown in figure \ref{fig:den-side-by-side} next to the true distribution. Visually, the extracted distribution is virtually identical to the true distribution, which is emphasized by the low uncertainties of $\sigma_\rho / \rho = \mathcal{O}(10^{-3})$ throughout the asteroid. Despite the low uncertainty, the density distribution is still consistent with the truth, with $\chi^2_r = 1.46$ using uncertainties calculated via section \ref{sec:density-uncertainty-model}. The number of asteroid points at which this uncertainty was calculated was about 6 million, making $\chi^2_r = 1.46$ actually quite high, due to the fact that our density uncertainty model does not take the correlation between density at different points into account. By comparison, the difference between the extracted distribution and a nearly-uniform distribution is $\chi^2_r = 6.4 \times 10^{6}$. The lumpy model's successful identification of a core is thus very statistically significant.

@@

To study the output of the density distributions, we start with executing them on synthetic data generated for the two reference asteroids. The result is shown in figure \ref{fig:den-uniform} displays density distributions and uncertainties. These distributions are consistent with the moments extracted by the MCMC in section \ref{sec:fit} --- indeed all the distributions we show have moments consistent with the data, even for non-ellipsoidal asteroids (not shown).

@@

We also explore model behaviour in non-uniform density asteroids. First we consider again the off-center core asteroid used in section \ref{sec:results-distro}; distributions are shown in figure \ref{fig:den-move}. As remarked before, the lumpy model fit is successful. However, the finite element model does not reproduce the core.The deviation from the true density extends to as much as 17\% in some locations, leading to a maximum significance of $0.52\sigma$. Visually the finite element model has spread the high-density core into the rest of the asteroid.

@@

The success of the model in the off-center case was due to the fact that the shape of the asteroid was offset from the center of mass by a corresponding amount, assumed to be known precisely. The position of the lump was therefore observed up to one free parameter: the lump's mass. In the centred core case, the core mass does not affect the asteroid center of mass so it is unconstrained, leading to large uncertainty. The underlying assumption that the asteroid's center of mass is so precisely known stems from the fact that the shape of the asteroid is observed to rotate around its center of mass. If observations do not allow the center of mass to be determined in this way, then the lump's position will be more uncertain for the off-center case.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figs/unc-scan.pdf}
  \caption{Density uncertainty as a function of physical / observational encounter properties (\textit{top / bottom}). Both the lumpy and finite element models are used, though the finite element model uncertainty is model-dominated and not very sensitive to encounter properties. The uncertainty threshold (\textit{red line}) is shown and the properties that exceed this threshold are shaded. The vertical black dotted lines are the property values of the reference asteroid. The figures illustrate the marked difference in density distributions that results from using different models.}
  \label{fig:net-uncertainty}
\end{figure*}

Figure \ref{fig:net-uncertainty} depicts $\langle \sigma_\rho / \rho \rangle$ as a function of these other encounter properties, except for the dependence of $\langle \sigma_\rho / \rho \rangle$ on initial spin direction which is better shown as a map in figure \ref{fig:spin-uncertainty}. We immediately see that the two density distribution models produce very different density uncertainties, due to the strong model-dependence of their output. For the most part, we consider the lumpy model's output because it is more sensitive to the encounter properties than the finite element model.

Figure \ref{fig:net-uncertainty} reveals that the most limiting physical properties of the asteroid are its perigee and period. Excess velocity does not greatly affect the uncertainty of the density distribution, and asteroid length is only important for $a_\mathcal{A}\sim $tens of meters. By contrast, the radius of Apophis is  $\sim 300$ m and its perigee and rotational period also obey the strong constraints \citep{giorgini2008predicting}. The comparison to Apophis is complicated by the fact that Apophis is smaller than our $a_\mathcal{A}$ value, is tumbling \citep{PRAVEC201448}, and may change slightly in physical properties due to tidal interaction during the encounter \citep{yu2014numerical,hirabayashi2021finite}. Further work must therefore be done to apply this analysis to Apophis.

@@

On the other hand, the cadence of observations $\Delta t$ and the presence of gaps in the data $T_\text{gap}$ do not affect results as strongly. Short cadence appears preferable, though as much as 20--40 min between observations is still not as detrimental to data quality as other properties. Likewise gaps appearing with size $T_\text{gap}$, even hours in length, do not greatly increase density uncertainty.

To quantify the information contained in \ref{fig:net-uncertainty}, we define a ``threshold'' on each encounter property beyond which we say the data quality is too low to extract meaningful information concerning the density distribution of the asteroid. The value of this threshold for perigee is set at the point when the finite element model produces $\langle \sigma_\rho / \rho \rangle = 30\%$, since this is about the maximum uncertainty that the lumpy model can produce. We cannot use this process to define the threshold for the other properties because the finite element model does not depend strongly enough on other properties, so we instead transfer the threshold definition to the value of $\langle \sigma_\rho / \rho \rangle$ produced by the lumpy model at that perigee ($5\times 10^{-4}$). This threshold is marked as a horizontal red line in figure \ref{fig:net-uncertainty}, and the values at which the encounter parameters exceed the threshold are shown in table \ref{tab:threshold-summary}.

@@

To address the dependence of $\langle \sigma_\rho / \rho \rangle$ on the choice of model, we also investigate the model-independent moment uncertainty $\sigma(K_{\ell m})$ as well, defined as the range of $K_{\ell m}$ values that contains 68.27\% of the marginal PPD for each $\ell$ and $m$. Since there is no degeneracy between moments and the actual encounter data, $\sigma(K_{\ell m})$ is well-defined and less noise-prone than $\langle \sigma_\rho / \rho \rangle$, though its physical relevance is not as obvious.

@@

By contrast, density moment uncertainty shows only a slight increase with $v_\infty$. This is likely due to the fact that larger $v_\infty$ leads to a faster and flatter orbit with less time spent close to the planet, where tidal torque is strongest. The change in encounter timing also adjusts the orientation of the asteroid at perigee, which has a separate effect on moment uncertainty. We control for this undesired orientation dependence by adjusting $\gamma_0$ so that, if no tidal torque is present, the orientation of the asteroid at perigee is constant for all orbits. The effect of tidal torque then causes orientations to be roughly the same at perigee where tidal torque is highest. This effectively removes the orientation-induced uncertainty.

\subsubsection{Rotational period}
\label{sec:scan-period}

In figure \ref{fig:scan-period}, we show $\sigma(K_{\ell m})$ as a function of rotational period $P_\omega$. As in the above section, the value of $\gamma_0$ was corrected to ensure roughly constant orientation at perigee. $K_{20}$ and $K_{22}$ show very large uncertainty for $P_\omega \lesssim 4$ hr because these fast rotators tumble very little after perigee. This increases uncertainty on the $K_{2m}$ parameters, which are largely constrained by post-encounter tumbling.

@@

\subsubsection{Initial spin pole}
\label{sec:scan-spin}

The tidal torque experienced by the asteroid is affected by the initial direction of asteroid spin $\bm \Omega_0$ both because spin sets the initial asteroid orientation up to $\gamma_0$ and because of the spin-dependence of the rotational equations of motion (equation \ref{eqn:omega-eom}).

@@

\subsubsection{Cadence}
\label{sec:scan-cadence}

>>>>

\begin{abstract}
  Knowledge of the interior density distribution of an asteroid can reveal its composition and constrain its evolutionary history. However, most asteroid observational techniques are not sensitive to interior properties. We investigate the interior constraints accessible through monitoring variations in angular velocity during a close encounter. We derive the equations of motion for a rigid asteroid's orientation and angular velocity to arbitrary order and use them to generate synthetic angular velocity data for a representative asteroid on a close Earth encounter. Using Markov Chain Monte Carlo fits, we perform injection-retrieval tests on these synthetic data to gain insights into the extent to which interior properties can be constrained. We also perform a sensitivity analysis of such an inversion technique to asteroid parameters (e.g., asteroid shape and orbital elements), observational set-up (e.g., measurement precision and cadence), and mapping models to convert constraints on the density moments to density distributions. We find that high precision in rotational period estimates ($\lesssim 50$ milliseconds) are necessary for each cadence, and that low perigees ($\lesssim 10$ Earth radii) are necessary to resolve density distributions.
\end{abstract}

@@

We begin our simulation at $D = 10 r_p$. Since the leading order of equation \ref{eqn:tidal-torque} yields $\tau \propto D^{-3}$, this corresponds roughly to a torque of $10^{-3}$ times the maximum torque at perigee. Unless otherwise indicated, the simulation is terminated at $D=10 r_p$ as well. With the simulation inputs specified, the equations of motion are integrated via the Runge-Kutta fourth order method, with a variable time step, manually chosen to limit integration error to only $\sim 100$ times the floating point numerical error.

@@

Before the MCMC was run, local minima in the likelihood were found via the Nelder-Mead algorithm implemented in \texttt{scipy} \citep{Gao2012}. Generally, only one local minimum existed, except when $K_{22}=0$ in which case rotational symmetry caused all values of $\gamma_0$ to be degenerate. Ensemble walkers were initialized near the local minimum, distributed by a Gaussian approximation of the likelihood as determined via the Hessian of the likelihood at the minimum. Due to the high sensitivity of the angular velocity data to density moments, the minimization procedure sometimes failed to isolate the minimum likelihood. Therefore, a simpler simulation without the $K_{3m}$ terms of equation \ref{eqn:tidal-torque} was first used to minimize likelihood as a function of the first-order parameters $\gamma_0$ and $K_{2m}$, and then the full simulation was used to find the second-order parameters $K_{3m}$ with the first-order parameters fixed.

@@

We assume that the asteroid's surface is known in the inertial frame from radar data. Since the asteroid tumbles during the encounter, we also assume that the center of mass of the asteroid is known in the inertial frame. The density moments are extracted from flyby data, but these are known instead in the body-fixed frame.

To compare the asteroid surface and extracted density moments in the same coordinate system, we define a new frame called the ``hybrid frame.'' The hybrid frame is co-located with the body-fixed frame, but its orientation is known with respect to the inertial frame; it has $\bm \Omega_0 \parallel \unit z_\text{hybrid}$ with third Euler angle $\gamma = 0$. The hybrid frame initially differs from the body-fixed frame only by a rotation around $\unit z_\text{hybrid}=\unit z_\text{body-fixed}$ of $\gamma_0$. Such rotations affect density moments via 
\begin{equation}
  K_{\ell m}^\mathrm{hybrid} = e^{-im\gamma_0}K_{\ell m}^\mathrm{body-fixed}.
  \label{eqn:body-fixed-to-hybrid}
\end{equation}
by equation \ref{eqn:ylm-rotation}. Thus, values and  uncertainties on $K_{\ell m}^\mathrm{body-fixed}$ and $\gamma_0$ (obtained from the encounter data) can be translated into values and uncertainties on $K_{\ell m}^\mathrm{hybrid}$. The surface model can be rotated from the inertial frame to the hybrid frame so that both the surface and $K_{\ell m}$ are now in the same frame and are comparable. Henceforth, we will operate only in the hybrid frame and suppress the label.

@@

To define the ``finite element'' model, we divide the asteroid into $N$ finite elements of uniform density and use the density $\rho_i$ of each as parameters. The four constraints imposed by the known mass and center of mass of the asteroid (seven with $K_{21}$ and $\Im K_{22}$ are also fixed) are used to fix some of the $\rho_i$. These constrained densities are easily computed since the asteroid mass $\mu_\mathcal{A}$ and the product $I_\mathcal{A}K_{\ell m}$ are both linear functions of $\rho_i$. Because $I_\mathcal{A}K_{00} = a_\mathcal{A}^2\mu_\mathcal{A}$ by definition (which is known), and the other fixed moments are zero, $I_\mathcal{A} K_{\ell m}$ is known for all the fixed moments, and computing the corresponding constrained densities is simply a matrix inversion. The size and location of each finite element must be chosen before extracting its density.

@@

The translation rules of spherical harmonics \citep{Gelderen1998TheSO} give that the asteroid density moments in the hybrid frame are
\begin{equation}
  \begin{aligned}
    K_{\ell m} = &\brackets{\sum_{i=0}^N \frac{a_i^2 + r_i^2}
    {a_\mathcal{A}^2}\mu_i}^{-1} \\
    &\times \brackets{\sum_{i=0}^N \sum_{\ell' m'}\mu_i
    \frac{a_i^{\ell'}}{a_\mathcal{A}^\ell}
    R_{\ell - \ell', m - m'}(\bm r_i)K_{\ell' m'}^{(i)}} \\
  \end{aligned}
\end{equation}
where the unmarked sum limits are $0 \leq l' \leq l$ and $-\ell' \leq m' \leq \ell'$. We also have total mass and center of mass constraints:
\begin{equation}
  \mu_\mathcal{A} = \sum_{i=0}^N \mu_i;  \qquad 0 = \sum_{i=0}^N \mu_i \bm r_i.
\end{equation}
Additional assumptions can be imposed on $K_{\ell m}^{(i)}$. For example, we can require that the lumps be ellipsoids, so that $K_{3 m}^{(i)} = 0$. The most extreme case is spherical lumps, which have $K_{\ell m} = 0$ for $\ell > 0$. $K_{00}=1$ is also guaranteed by definition, meaning that each spherical lump has only five DOF ($a_i$, $\mu_i$, and $\bm r_i$). The substrate has one degree of freedom, since only $\mu_0$ is unknown. Thus, this spherical lumpy model has $5N - 3$ total DOF. Again, the choice of $N$ affects the accuracy and uncertainty of the model results. For the rest of this paper, we use $N=1$ spherical lump for simplicity, corresponding to 2 DOF. In section \ref{sec:density-compare}, we also use $N=2$ spherical lumps which possesses 7 DOF.

@@

Once the parameters of the density distribution models have been extracted, each MCMC sample corresponds to its own density distribution. To generate an average distribution, we randomly choose 1,000 of these samples and define the density distribution $\rho$ at each point to be the mean of the densities at that point across the 1,000 samples, and take the standard deviation of the sample densities as the (uncorrelated) density uncertainty $\sigma_\rho$. In the case of the finite element model, we run the MCMC twenty times and select 1,000 parameter samples from each to reduce dependence on the initial choice of finite element locations.

@@

\begin{enumerate}
  \item An orbit around a spherical, non-rotating, Moonless Earth with $6$ km s$^{-1}$ excess velocity and perigee at 5 Earth radii. This orbit was chosen to roughly match that of Apophis and corresponds to an eccentricity of 3.88. 
  \item An initial roll of $\gamma_0=\pi/8$.
  \item A cadence of 2 minutes and observational uncertainty of $\sigma_\theta = 10^{-2}$ and $\sigma_P / P = 10^{-7}$.
  \item A rotational period of 9 hours, with the initial angular velocity vector distributed between the $\unit X$, $\unit Y$, and $\unit Z$ axes in a $1:2:-2$ ratio.
  \item An asteroid with radius $a_\mathcal{A} = 1$ km and $K_{3m}=0$. For $K_{22}$ and $K_{20}$, we use two standard values: one with $(K_{22}, K_{20}) = (0, -0.097)$ and one with $(0.052, -0.202)$. Including the third point obtained by reflection $K_{22}\rightarrow -K_{22}$ (a 90$^\circ$ rotation), these are the three points that minimize the mean distance between an arbitrary point in the allowed parameter space (equation \ref{eqn:parameter-bounds}) and these reference values. The first point is called the symmetric case because the corresponding uniform-density-ellipsoid model is rotationally symmetric around $\unit z$. The second case is called the asymmetric case. The asymmetric case has $a=1140$ m, $b=1839$ m, and $c=565$ m, while the symmetric case has $a=b=1411$ m and $c=1008$ m.
\end{enumerate}

In the following sections, we first introduce the retrieval capabilities regarding density moments, then we turn to an example of a retrieval of the density distribution from the moments. For both stages of information retrieval, we find that the results are consistent with the synthetic data and the true asteroid interior.

\subsection{Density moment retrieval}
\label{sec:results-moment}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/example-residuals.pdf}
  \caption{Data, best-fitting results, and residuals for a fit to synthetic data simulated for the asymmetric reference asteroid. Uncertainty bands are also shown. The best fit results are consistent with the data.}
  \label{fig:example-residuals}
\end{figure}

@@

Figure \ref{fig:example-corner} shows a corner plot of the PPDs of the ten parameters (namely $\gamma_0$ and $K_{\ell m}$ for $\ell \leq 3$), marginalized to functions of one (histograms) or two (contours) variables. The true parameters are shown and usually lie within 1 or 2$\sigma$ of the $\Delta K_{\ell m} = 0$, where $\Delta K_{\ell m}$ is the difference between the mean posterior $K_{\ell m}$ and the true $K_{\ell m}$. The PPDs are generally Gaussian and sometimes show strong correlation between parameters, but no continuous degeneracy occurs. We performed 48 independent minimizations of the likelihood before the MCMC fit began, each with an initial point chosen randomly in the parameter space. All converged to the same minimum, demonstrating that the model lacks discrete degeneracy as well.

@@

To provide an example of a density distribution extraction, we consider an asteroid with a core and ask whether our methodology can resolve the location, mass, and size of the core. Specifically, we use a spherical core of radius 300 m, placed 500 m from the center and with density 50\% greater than that of the surrounding asteroid. We use the lumpy model to extract a distribution in this section because it is designed to look for cores.

Synthetic data was generated for the new asteroid and moments were extracted via the process described above. These moments were again consistent with the true moments, with a $\chi^2$ per DOF of $\chi^2_r = 1.1$, and inconsistent with the moments of a nearly-uniform asteroid of the same shape ($\chi^2_r = 2.7 \times 10^8$). The inconsistency comes almost entirely from the MOI-constraining $K_{2m}$ moments, which are sensitive to large mass distribution changes such as the presence of a core.

@@

  \includegraphics[width=0.49\linewidth]{figs/side-by-side-u}
  \caption{Cross-sectional slices of the extracted distribution (\textit{left}) and true distribution (\textit{right}) of the asymmetric reference asteroid. The density distribution was created via the lumpy model. Uncertainty in the distribution is shown below. The extracted distribution is nearly identical to the true distribution and is statistically consistent with the truth.}
  \label{fig:den-side-by-side}
\end{figure}

A density distribution was then extracted, shown in figure \ref{fig:den-side-by-side} next to the true distribution. Visually, the extracted distribution is virtually identical to the true distribution, which is emphasized by the low uncertainties of $\sigma_\rho / \rho = \mathcal{O}(10^{-3})$ throughout the asteroid. Despite the low uncertainty, the density distribution is still consistent with the truth, with $\chi^2_r = 1.46$ using uncertainties calculated via section \ref{sec:density-uncertainty-model}. The number of asteroid points at which the $\chi^2$ was evaluated was about 6 million, making $\chi^2_r = 1.46$ actually quite high. This is due to the fact that our density uncertainty model does not take the correlation between density at different points into account. By comparison, the difference between the extracted distribution and a nearly-uniform distribution is $\chi^2_r = 6.4 \times 10^{6}$. The lumpy model's successful identification of a core is thus statistically significant.

@@

To study the output of the density distribution models, we begin by executing them on synthetic data generated for the two reference asteroids. The resulting density distributions and uncertainties are shown in figure \ref{fig:den-uniform}. These distributions are consistent with the moments extracted by the MCMC in section \ref{sec:fit} --- indeed all the distributions we show have moments consistent with the data, even for non-ellipsoidal asteroids (not shown).

@@

We also explore model behaviour in non-uniform density asteroids. First we consider again the off-center core asteroid used in section \ref{sec:results-distro}; distributions are shown in figure \ref{fig:den-move}. As remarked before, the lumpy model fit is successful. However, the finite element model does not reproduce the core. The deviation from the true density extends to as much as 17\% in some locations, leading to a maximum significance of $0.52\sigma$. Visually the finite element model has spread the high-density core into the rest of the asteroid.

@@

The success of the model in the off-center case was due to the fact that the shape of the asteroid was offset from the center of mass by a corresponding amount, assumed to be known precisely. The position of the lump was therefore observed up to one free parameter: the lump's mass. In the centred core case, the core mass does not affect the asteroid center of mass so the mass is unconstrained and uncertain. The underlying assumption that the asteroid's center of mass is so precisely known stems from the fact that the shape of the asteroid is observed to rotate around its center of mass. If observations do not allow the center of mass to be determined in this way, then the lump's position will be more uncertain for the off-center case.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figs/unc-scan.pdf}
  \caption{Density uncertainty as a function of physical / observational encounter properties (\textit{top / bottom}). Both the lumpy and finite element models are used, though the finite element model uncertainty is model-dominated and not very sensitive to encounter properties. The uncertainty threshold (\textit{red line}) is shown and the properties that exceed this threshold are shaded. The vertical black dotted lines are the property values of the reference asteroid. Perigee, rotational period, and observational uncertainty most strongly affect density uncertainty.}
  \label{fig:net-uncertainty}
\end{figure*}

Figure \ref{fig:net-uncertainty} depicts $\langle \sigma_\rho / \rho \rangle$ as a function of these other encounter properties, except for the dependence of $\langle \sigma_\rho / \rho \rangle$ on initial spin direction which is better shown as a map in figure \ref{fig:spin-uncertainty}. We immediately see that the two density distribution models produce very different density uncertainties, due to the strong model-dependence of their output. We focus on the lumpy model's output because it is more sensitive to the encounter properties than the finite element model.

Figure \ref{fig:net-uncertainty} reveals that the most limiting physical properties of the asteroid are its perigee and period. Excess velocity does not greatly affect the uncertainty of the density distribution, and asteroid length is only important for $a_\mathcal{A}\sim$ tens of meters. By contrast, the radius of Apophis is  $\sim 300$ m and its perigee and rotational period also obey the strong constraints \citep{giorgini2008predicting}. The comparison to Apophis is complicated by the fact that Apophis is smaller than our $a_\mathcal{A}$ value, is tumbling \citep{PRAVEC201448}, and may change slightly in physical properties due to tidal interaction during the encounter \citep{yu2014numerical,hirabayashi2021finite}. Further work must therefore be done to apply this analysis to Apophis.

@@

On the other hand, the cadence of observations $\Delta t$ and the presence of gaps in the data $T_\text{gap}$ do not affect results as strongly. Short cadence appears preferable, though as much as 20--40 min between observations is still not as detrimental to data quality as other properties. Likewise, gaps appearing with size $T_\text{gap}$, even hours in length, do not greatly increase density uncertainty.

To summarize the information contained in \ref{fig:net-uncertainty}, we define a ``threshold'' on each encounter property beyond which we say the data quality is too low to extract meaningful information concerning the density distribution of the asteroid. The value of this threshold for perigee is set at the point when the finite element model produces $\langle \sigma_\rho / \rho \rangle = 30\%$, since this is just under the maximum uncertainty that the lumpy model produces. We cannot use this process to define the threshold for the other properties because the finite element model does not depend strongly enough on other properties, so we instead transfer the threshold definition to the value of $\langle \sigma_\rho / \rho \rangle$ produced by the lumpy model at that perigee value ($\langle \sigma_\rho / \rho \rangle = 5\times 10^{-4}$). This threshold is marked as a horizontal red line in figure \ref{fig:net-uncertainty}, and the values at which the encounter parameters exceed the threshold are shown in table \ref{tab:threshold-summary}.

@@

To address the dependence of $\langle \sigma_\rho / \rho \rangle$ on the choice of model, we also investigate the model-independent moment uncertainty $\sigma(K_{\ell m})$ as well, defined as the range of $K_{\ell m}$ values that contains 68.27\% of the marginal PPD. Since there is no degeneracy between moments and the actual encounter data, $\sigma(K_{\ell m})$ is well-defined and less noise-prone than $\langle \sigma_\rho / \rho \rangle$, though its physical relevance is not as obvious.

@@

By contrast, density moment uncertainty shows only a slight increase with $v_\infty$. This is likely due to the fact that larger $v_\infty$ leads to a faster and flatter orbit with less time spent close to the planet, where tidal torque is strongest. The change in encounter timing also adjusts the orientation of the asteroid at perigee, which has a separate effect on moment uncertainty. We correct this undesired orientation dependence by adjusting $\gamma_0$ so that $\gamma$ is roughly the same value at perigee (where torque is highest) for all points in the data set.

\subsubsection{Initial angular velocity}
\label{sec:scan-period}

In figure \ref{fig:scan-period}, we show $\sigma(K_{\ell m})$ as a function of initial rotational period $P_\omega$. As in the above section, the value of $\gamma_0$ was corrected to ensure roughly constant orientation at perigee. $K_{20}$ and $K_{22}$ show very large uncertainty for $P_\omega \lesssim 4$ hr because these fast rotators tumble very little after perigee. This increases uncertainty on the $K_{2m}$ parameters, which are largely constrained by post-encounter tumbling.

@@

The tidal torque experienced by the asteroid is also affected by the initial direction of asteroid spin $\bm \Omega_0$ both because spin sets the initial asteroid orientation up to $\gamma_0$ and because of the spin-dependence of the rotational equations of motion (equation \ref{eqn:omega-eom}).

@@

\subsubsection{Cadence and data gaps}
\label{sec:scan-cadence}
